---
title: 'Week 4: home assignment'
author: "Manuel Trombetta"
date: "2026-02-05"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(message = FALSE)
knitr::opts_chunk$set(warning = FALSE)
knitr::opts_chunk$set(error = FALSE)
```

### Load libraries

```{r}
library(tidyverse)
library(broom)
library(car)
library(languageR)
```

### Load data from "languageR" library

```{r}
data("dative")
head(dative)
```

### Formulating null hypotheses

- **H0~1~ (Animacy)**  
  *The animacy of the recipient does not affect the choice between PP and NP.*

- **H0~2~ (Length)**  
  *The length of the recipient does not affect the choice between PP and NP.*
  
### Creation of the logistic model

First, the dataset was loaded and the response variable (RealizationOfRecipient) was converted into a factor. 
In R, a factor is used to represent categorical variables; in this case, the response has two possible categories (PP and NP). 
A logistic regression model with a logit link was then fitted, using animacy and length of the recipient as predictors. 
This model estimates how these factors affect the probability of selecting the PP construction.


```{r}
# using factor on RealizationOfRecipient:
dative <- dative %>%
  mutate(RealizationOfRecipient = factor(RealizationOfRecipient))

dative_glm <- glm(RealizationOfRecipient ~ AnimacyOfRec + LengthOfRecipient,
                  data = dative,
                  family="binomial")

vif(dative_glm)
summary(dative_glm)
tidy(dative_glm)
```
### Interpretation of the results

Both predictors show significant effects.  
Animacy has a positive and significant effect: inanimate recipients are more likely to occur in the PP construction (Estimate = 1.02, SE = 0.15, p < 2.28e-11).  
Length of the recipient also has a strong positive effect (Estimate = 0.68, SE = 0.04, p < 2e-16), indicating that longer recipients favor the PP construction.

Both null hypotheses are therefore rejected (the p-value is less than 0.05).
Keeping animacy constant, the probability of the PP construction increases substantially with recipient length.



### Assessment of the probablity of the *PP category*

```{r}
intercept <- coef(dative_glm)[[1]]
slope_length <- coef(dative_glm)[[3]]
slope_anim <- coef(dative_glm)[[2]]

plogis(intercept + slope_length*5 + slope_anim*0) # 5-word recipient
plogis(intercept + slope_length*3 + slope_anim*0) # 3-word recipient
plogis(intercept + slope_length*1 + slope_anim*0) # 1-word recipient
```
For a five-word recipient, the model predicts a high probability of the PP construction (about 73%), even when the recipient is animate.As for 3-word recipients and 1-word recipients, the probabilities drop to 41% and 15% respectively.


### EXTRA: Visualisation of predicted probabilities

```{r}
data_anim <- data.frame(
  LengthOfRecipient = seq(1,6, by=1),
  AnimacyOfRec = "animate"
)

data_inanim <- data.frame(
  LengthOfRecipient = seq(1,6, by=1),
  AnimacyOfRec = "inanimate"
)

probabilities_anim <- predict(
  dative_glm,
  newdata = data_anim,
  type = "response"
)

probabilities_inanim <- predict(
  dative_glm,
  newdata = data_inanim,
  type = "response"
)

#check
head(probabilities_anim)
head(probabilities_inanim)
```
```{r}
# Visualisation of predicted probabilities for animate and inanimate recipients.
plot(
  data_anim$LengthOfRecipient,
  probabilities_anim,
  type = "l",
  lwd = 2,
  col = "red",
  xlab = "Length of recipient",
  ylab = "Predicted probability of PP construction",
  ylim = c(0, 1)
)

lines(
  data_inanim$LengthOfRecipient,
  probabilities_inanim,
  lwd = 2,
  col = "blue"
)

legend(
  "topleft",
  legend = c("Animate", "Inanimate"),
  col = c("red", "blue"),
  lwd = 2
)
```

The curves illustrate the predictions of the logistic regression model, highlighting a positive effect of recipient length and a strong effect of animacy on the choice of the PP construction.


source: https://www.r-bloggers.com/2023/10/plotting-a-logistic-regression-in-base-r/
