---
title: 'Week 5: home assignment'
author: "Manuel Trombetta"
date: "2026-02-14"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(message = FALSE)
knitr::opts_chunk$set(warning = FALSE)
knitr::opts_chunk$set(error = TRUE)
```

### Mixed-Effects Analysis of Visual Lexical Decision Times


# Task A - **Load Packages and dataset:**

```{r}
library(tidyverse)
library(lme4)
library(MuMIn)
library(tidyr)
library(broom)
library(ggeffects)
library(languageR)
```

look at the data:

```{r}
data <- lexdec 
head(data)
```

# Task B - **Filter the data:**
```{r}
data <- as_tibble(data) %>%
  filter(RT < 7, Correct == "correct") 

# count rows:
nrow(data)
```

# Task C - **Creation of the mixed linear regression model:**
```{r}
data_lmer <- lmer(RT ~ Trial + Length +
                    (1|Subject) + # random intercept
                    (1|Word),
                  data = data,
                  REML = FALSE)
summary(data_lmer)
```

We include random intercepts for Subject and Word because the data contain repeated measurements: *each subject responds to many words and each word is seen by many subjects*. Participants differ in their overall speed, and words differ in their difficulty.

Looking at the variance components, the variance for Subject (0.0177) is close to the residual variance (0.0225), indicating **substantial differences in baseline reaction times across participants**. In contrast, the variance for Word (0.0034) is much smaller relative to the residual variance, suggesting that variability across lexical items is present but less pronounced.

# Task D - **Significance testing of predictors**
```{r}
# model with no Trial:
model_noTrial <- lmer(RT ~ Length +
                        (1|Subject) + (1|Word),
                      data = data,
                      REML = FALSE)

anova(model_noTrial, data_lmer)

# model with no Length:
model_noLength <- lmer(RT ~ Trial +
                         (1|Subject) + (1|Word),
                       data = data,
                       REML = FALSE)

anova(model_noLength, data_lmer)
```
The **chi-square statistic** indicates how much the model improves when a predictor is added, the **degrees of freedom** represent the number of added parameters, and the **p-value** shows whether this improvement is statistically significant.

Adding Trial significantly improves the model (Chisq = 5.03, df = 1, p = .025), showing that reaction times change during the experiment. Adding Length improves the model even more (Chisq = 19.04, df = 1, p < .001), indicating that longer words lead to longer reaction times. Both null hypotheses are rejected.


# Task E - **Subject variation in reaction time across Trials**

Add random slope:
```{r}
data_lmer2 <- lmer(RT ~ Trial + Length +
                    (1 + Trial|Subject) + # random slope
                    (1|Word),
                  data = data,
                  REML = FALSE)
summary(data_lmer2)
```
Create predictions for Subject:
```{r}
mymodel_pred <- predict_response(
  data_lmer2,
  terms = c("Trial", "Subject"),
  type = "random")

```

Create the plot:
```{r}
plot(mymodel_pred, show_ci = FALSE) +
  scale_color_manual(values = rainbow(21))

```

**Interpretation**:

Most subjects show decreasing predicted reaction times across trials, suggesting a practice effect. A few participants display flatter or slightly increasing trends, indicating individual variability and possible fatigue.
